Discrete Quantum Theories and Computing

### 2018-11-23 dissertation.tex
dissertation.lyx "Chapter 2 Conventional Quantum Theory" Updating...

The part of conventional quantum theory (CQT) used by quantum circuit
model is described by the following:
\begin{enumerate}[label=(\roman{enumi})]
\item \label{enu:-orthonormal-basis}$D$ orthonormal basis vectors for
a Hilbert space of dimension~$D$, 
\item \label{enu:probability-amplitude}$D$ complex probability amplitude
coefficients describing the contribution of each basis vector, 
\item \label{enu:unitary-matrix}a set of probability-conserving unitary
matrix operators that suffice to describe all required state transformations
of a quantum circuit, 
\item \label{enu:measurement-framework}and a measurement framework. 
\end{enumerate}
In Sec.~\ref{CQC1qubitBloch.sec}, we focus on the geometric issues
raised by the properties \ref{enu:-orthonormal-basis} and \ref{enu:probability-amplitude}
given above for CQT. In Sec.~\ref{sec:fuzzy}, we introduce the important
issues of \ref{enu:measurement-framework} and the foundations of
quantum probability space. The observables and expectation values
defined in Sec.~\ref{sec:fuzzy} will be used to understand the geometry
of entangled states in Sec.~\ref{sec:The-geometry-of}. The property~\ref{enu:unitary-matrix}
directly related to quantum circuit will be introduced later in Chapter~\ref{chap:DQT=000026DQC}
when we dive into quantum algorithms.

@@

A \emph{probability space} is a mathematical abstraction specifying
the necessary conditions for reasoning coherently about collections
of uncertain events \cite{Kolmogorov1950,544199,Griffiths2003,Grabisch2016}.
Given a finite sample space~$\Omega$ representing all possible outcomes
of a process, and its power set $2^{\Omega}$ as the classical event
space, a classical probability measure~$\mu$ maps an event~$E$
to a number $\mu\left(E\right)\in\left[0,1\right]$ specifying how
likely one of the outcome in $E$ actually happens. To maintain the
coherence, $\mu$ is subject to the following constraints: $\mu\left(\emptyset\right)=0$,
$\mu\left(\Omega\right)=1$, and $\mu\left(\overline{E}\right)=1-\mu\left(E\right)$,
where $\overline{E}$ is the complement of $E$.  Moreover, we require
$\mu\left(E_{0}\cup E_{1}\right)=\mu\left(E_{0}\right)+\mu\left(E_{1}\right)$
for each pair of disjoint events $E_{0}\subseteq\Omega$ and $E_{1}\subseteq\Omega$.

Since the previous abstraction doesn't specify the process to generate
the outcomes, this process could well be a quantum experiment. Let
us prepare a beam of one kind of spin~$1$ particles with basis vectors
$\ket{0}$, $\ket{1}$, and $\ket{2}$. In principle, this beam can
be split by a Stern-Gerlach type experiment according to the eigenvalues
of the observable $\mathbf{O}_{0}$ with with spectral decomposition
\begin{equation}
\mathbf{O}_{0}=0\,\proj{0}+1\,\proj{1}+2\,\proj{2}\,,\label{eq:observable}
\end{equation}
and the states corresponding to the split beams are $\ket{0}$, $\ket{1}$,
and $\ket{2}$ \cite{:/content/aip/journal/jmp/21/1/10.1063/1.524312,peres1995quantum}.
Instead of sending a beam, if only one particle is sent to the beam
splitter, the state after the experiment is one of $\ket{0}$, $\ket{1}$,
and $\ket{2}$, and there is a probability for each post-experimental
state. In the language of our abstraction, the set of outcomes is
$\Omega_{0}=\left\{ \ket{0},\ket{1},\ket{2}\right\} $. All possible
events are $\emptyset$, $\left\{ \ket{0}\right\} $, $\left\{ \ket{1}\right\} $,
$\left\{ \ket{2}\right\} $, $\left\{ \ket{0},\ket{1}\right\} $,
$\left\{ \ket{0},\ket{2}\right\} $, $\left\{ \ket{1},\ket{2}\right\} $,
and $\Omega_{0}$. And this experiment defines a classical probability
measure $\mu_{0}:2^{\Omega_{0}}\rightarrow\left[0,1\right]$.

One special feature of quantum experiments is that the probabilities
in different experiments are correlated. Consider another experiment
similar to the previous one. While sending exactly the same particle,
we use a different beam splitter corresponding to the observable $\mathbf{O}_{1}$
with spectral decomposition $\mathbf{O}_{1}=0\,\proj{\ps}+1\,\proj{\ms}+2\,\proj{2}$,
where 
\begin{align}
\ket{\ps}=\frac{\ket{0}+\ket{1}}{\sqrt{2}}\,, &  & \ket{\ms}=\frac{\ket{0}-\ket{1}}{\sqrt{2}}\,.\label{eq:psms}
\end{align}
Although the sample space $\Omega_{1}=\left\{ \ket{\ps},\ket{\ms},\ket{2}\right\} $
and the probability measure $\mu_{1}:2^{\Omega_{1}}\rightarrow\left[0,1\right]$
defined by this experiment is different from the previous one, these
two experiments may produce the same post-experimental state~$\ket{2}$,
and the probability of the common event $\left\{ \ket{2}\right\} $
is believed to be the same, i.e., $\mu_{0}\left(\left\{ \ket{2}\right\} \right)=\mu_{1}\left(\left\{ \ket{2}\right\} \right)$.
The general situation can be summarized as the following postulate.

\begin{postulate}Given two quantum experiments, their post-experimental
states are two orthonormal bases, and $E$ is a event for both of
the experiments. Assume we send the same particle into these two experiments,
then the probability of $E$ in these two experiments are the same.\end{postulate}

\noindent This postulate is actually equivalent to the fact that commuting
observables could be measured simultaneously. The latter fact is essential
to define contextuality which will be explained later in Sec.~\ref{sec:Kochen-Specker}.

Since the probability induced by the different beam splitters are
correlated, it is more natural to define one quantum event space~$\events$
containing all possible classical event spaces using different beam
splitters. However, simply taking the union of all event spaces is
bad because two events might appear different but representing the
same situation. For example, if we take the complement on the both
sides of $\mu_{0}\left(\left\{ \ket{2}\right\} \right)=\mu_{1}\left(\left\{ \ket{2}\right\} \right)$,
we will have
\begin{equation}
\mu_{0}\left(\left\{ \ket{0},\ket{1}\right\} \right)=1-\mu_{0}\left(\left\{ \ket{2}\right\} \right)=1-\mu_{1}\left(\left\{ \ket{2}\right\} \right)=\mu_{1}\left(\left\{ \ket{\ps},\ket{\ms}\right\} \right)\,,
\end{equation}
that is, the probabilities of events $\left\{ \ket{0},\ket{1}\right\} $
and $\left\{ \ket{\ps},\ket{\ms}\right\} $ are always the same. To
simplify our discussion, we want to identify two classical events
$\left\{ \ket{0},\ket{1}\right\} $ and $\left\{ \ket{\ps},\ket{\ms}\right\} $
as the same quantum event. Since the projectors generated by these
two events are equal as operators, i.e., $\proj{0}+\proj{1}=\proj{\ps}+\proj{\ms}$,
given any classical event~$E$, we can follow the same idea to define
its quantum event as the projector generated by $E$
\begin{equation}
\varphi\left(E\right)=\sum_{\ket{j}\in E}\proj{j}\label{eq:pullback-function}
\end{equation}
with the convention $\varphi\left(\emptyset\right)=\mathbb{0}$. This
function $\varphi$ well glues all classical event spaces into one
quantum event space~$\events$, the set of all projectors on a given
Hilbert space, because of the following lemma.

\begin{lemma}Given two quantum experiments, their post-experimental
states are two orthonormal bases $\Omega$ and $\Omega'$, and $E\subseteq\Omega$
and $E'\subseteq\Omega'$ are two events with $\varphi\left(E\right)=\varphi\left(E'\right)$.
Assume we send the same particle into these two experiments, then
the probability of $E$ is the same as the probability of $E'$.\end{lemma}

\begin{proof}This lemma is obvious if we consider the third orthonormal
bases $\Omega''=E\cup\left(\Omega'\backslash E'\right)$ and induction
on the basis vectors.\end{proof}

\noindent Moreover, this function $\varphi$ also naturally sends
the set structure to the corresponding projector structure: $\varphi\left(\Omega\right)=\mathbb{1}$
and $\varphi\left(\overline{E}\right)=\mathbb{1}-\varphi\left(E\right)$.
Given a sample space $\Omega$ and a pair of event $E_{0}\subseteq\Omega$
and $E_{1}\subseteq\Omega$, their corresponding quantum events $\varphi\left(E_{0}\right)$
and $\varphi\left(E_{1}\right)$ are commuting and vice versa, where
two projectors $P_{0}$ and~$P_{1}$ are \emph{commuting} if $P_{0}P_{1}=P_{1}P_{0}$.
Moreover, the events $E_{0}$ and $E_{1}$ in the same sample space
also satisfy the following properties: 
\begin{align}
\varphi\left(E_{0}\cap E_{1}\right) & =\varphi\left(E_{0}\right)\varphi\left(E_{1}\right)\,, & \varphi\left(E_{0}\cup E_{1}\right) & =\varphi\left(E_{0}\right)+\varphi\left(E_{1}\right)-\varphi\left(E_{0}\right)\varphi\left(E_{1}\right)\,.\label{eq:pullback-properties}
\end{align}
If $E_{0}$ and $E_{1}$ are disjoint, then $\varphi\left(E_{0}\right)$
and $\varphi\left(E_{1}\right)$ are orthogonal and vice versa, where
two projectors $P_{0}$ and $P_{1}$ are called \emph{orthogonal}
if $P_{0}P_{1}=\mathbb{0}$.

>>>>

The part of conventional quantum theory (CQT) used by quantum circuit
model is described by the following:
\begin{enumerate}[label=(\roman{enumi})]
\item \label{enu:-orthonormal-basis}$D$ orthonormal basis vectors for
a Hilbert space of dimension~$D$, 
\item \label{enu:probability-amplitude}$D$ complex probability amplitude
coefficients describing the contribution of each basis vector, 
\item \label{enu:unitary-matrix}a set of probability-conserving unitary
matrix operators that suffice to describe all required state transformations
of a quantum circuit, 
\item \label{enu:measurement-framework}and a measurement framework. 
\end{enumerate}
In Sec.~\ref{CQC1qubitBloch.sec}, we focus on the geometric issues
raised by the properties \ref{enu:-orthonormal-basis} and \ref{enu:probability-amplitude}
given above for CQT. In Sec.~\ref{sec:fuzzy}, we introduce the important
issues of \ref{enu:measurement-framework} and the foundations of
quantum probability space. The expectation values defined in Sec.~\ref{sec:fuzzy}
will be used to understand the geometry of entangled states in Sec.~\ref{sec:The-geometry-of}.
The property~\ref{enu:unitary-matrix} directly related to quantum
circuits will be introduced later in Chapter~\ref{chap:DQT=000026DQC}
when we dive into quantum algorithms.

@@

A \emph{probability space} is a mathematical abstraction specifying
the necessary conditions for reasoning coherently about collections
of uncertain events \cite{Kolmogorov1950,544199,Griffiths2003,Grabisch2016}.
Given a finite sample space~$\Omega$ representing all possible outcomes
of a process, and its power set $2^{\Omega}$ as the classical event
space, a classical probability measure~$\mu$ maps an event~$E$
to a number $\mu\left(E\right)\in\left[0,1\right]$ specifying how
likely one of the outcome in $E$ will actually happen. To maintain
the coherence, $\mu$ is subject to the following constraints: $\mu\left(\emptyset\right)=0$,
$\mu\left(\Omega\right)=1$, and $\mu\left(\overline{E}\right)=1-\mu\left(E\right)$,
where $\overline{E}$ is the complement of $E$.  Moreover, we require
$\mu\left(E_{0}\cup E_{1}\right)=\mu\left(E_{0}\right)+\mu\left(E_{1}\right)$
for each pair of disjoint events $E_{0}\subseteq\Omega$ and $E_{1}\subseteq\Omega$.

Since the previous abstraction doesn't specify the process to generate
the outcomes, this process could well be a quantum experiment. Let
us prepare a beam of one kind of spin~$1$ particles whose state
can be characterized by a vector in three-dimensional Hilbert space
with basis vectors $\ket{0}$, $\ket{1}$, and $\ket{2}$. In principle,
this beam can be split by a Stern-Gerlach type experiment according
to the eigenvalues of the observable $\mathbf{O}_{0}$ with spectral
decomposition 
\begin{equation}
\mathbf{O}_{0}=0\,\proj{0}+1\,\proj{1}+2\,\proj{2}\,,\label{eq:observable}
\end{equation}
and the states corresponding to the split beams are $\ket{0}$, $\ket{1}$,
and $\ket{2}$ \cite{:/content/aip/journal/jmp/21/1/10.1063/1.524312,peres1995quantum}.
Instead of sending a beam of particles, if only one particle is sent
to the beam splitter, the state of the particle after the experiment
is one of $\ket{0}$, $\ket{1}$, and $\ket{2}$, and there is a probability
to get each post-experimental state. In the language of our abstraction,
the set of outcomes is $\Omega_{0}=\left\{ \ket{0},\ket{1},\ket{2}\right\} $.
All possible events are $\emptyset$, $\left\{ \ket{0}\right\} $,
$\left\{ \ket{1}\right\} $, $\left\{ \ket{2}\right\} $, $\left\{ \ket{0},\ket{1}\right\} $,
$\left\{ \ket{0},\ket{2}\right\} $, $\left\{ \ket{1},\ket{2}\right\} $,
and $\Omega_{0}$. And this experiment defines a classical probability
measure $\mu_{0}:2^{\Omega_{0}}\rightarrow\left[0,1\right]$.

One special feature of quantum experiments is that the probabilities
in different experiments are correlated. Consider another experiment
sending exactly the same particle as the previous one, but using a
different beam splitter corresponding to the observable $\mathbf{O}_{1}$
with spectral decomposition $\mathbf{O}_{1}=0\,\proj{\ps}+1\,\proj{\ms}+2\,\proj{2}$,
where 
\begin{align}
\ket{\ps}=\frac{\ket{0}+\ket{1}}{\sqrt{2}}\,, &  & \ket{\ms}=\frac{\ket{0}-\ket{1}}{\sqrt{2}}\,.\label{eq:psms}
\end{align}
Although the sample space $\Omega_{1}=\left\{ \ket{\ps},\ket{\ms},\ket{2}\right\} $
and the probability measure $\mu_{1}:2^{\Omega_{1}}\rightarrow\left[0,1\right]$
defined by this experiment is different from the previous one, these
two experiments may produce the same post-experimental state~$\ket{2}$,
and the probability of the common event $\left\{ \ket{2}\right\} $
is believed to be the same, i.e., $\mu_{0}\left(\left\{ \ket{2}\right\} \right)=\mu_{1}\left(\left\{ \ket{2}\right\} \right)$.
In general, as long as sending the same particle, the probability
of the same event in different experiments should always be the same.
This fact is equivalent to the fact that commuting observables could
be measured simultaneously which is essential to define contextuality
and will be explained in Sec.~\ref{sec:Kochen-Specker}.

Since the probability induced by the different beam splitters are
correlated, it is more natural to define one quantum event space~$\events$
containing all possible classical event spaces using different beam
splitters. However, simply taking the union of all event spaces is
a bad idea because two events might appear different but representing
the same situation. For example, if we take the complement on the
both sides of $\mu_{0}\left(\left\{ \ket{2}\right\} \right)=\mu_{1}\left(\left\{ \ket{2}\right\} \right)$,
we will have
\begin{equation}
\mu_{0}\left(\left\{ \ket{0},\ket{1}\right\} \right)=1-\mu_{0}\left(\left\{ \ket{2}\right\} \right)=1-\mu_{1}\left(\left\{ \ket{2}\right\} \right)=\mu_{1}\left(\left\{ \ket{\ps},\ket{\ms}\right\} \right)\,,
\end{equation}
that is, the probabilities of events $\left\{ \ket{0},\ket{1}\right\} $
and $\left\{ \ket{\ps},\ket{\ms}\right\} $ are always the same, and
should be identified as the same quantum event to simplify our discussion.
Since the projectors generated by these two events are equal as operators,
i.e., $\proj{0}+\proj{1}=\proj{\ps}+\proj{\ms}$, given any classical
event~$E$, we can follow the same idea to define its quantum event
as the projector generated by $E$
\begin{equation}
\varphi\left(E\right)=\sum_{\ket{j}\in E}\proj{j}\label{eq:pullback-function}
\end{equation}
with the convention $\varphi\left(\emptyset\right)=\mathbb{0}$. This
function $\varphi$ well glues all classical event spaces into one
quantum event space~$\events$, the set of all projectors on a given
Hilbert space, because it is easy to inductively prove that if two
classical events~$E$ and $E'$ corresponding to the same quantum
event, i.e., $\varphi\left(E\right)=\varphi\left(E'\right)$, then
the probability of $E$ is the same as the probability of $E'$.

Moreover, this function $\varphi$ also naturally sends the set structure
to the corresponding projector structure: $\varphi\left(\Omega\right)=\mathbb{1}$
and $\varphi\left(\overline{E}\right)=\mathbb{1}-\varphi\left(E\right)$.
Given a sample space $\Omega$ and a pair of event $E_{0}\subseteq\Omega$
and $E_{1}\subseteq\Omega$, their corresponding quantum events $\varphi\left(E_{0}\right)$
and $\varphi\left(E_{1}\right)$ are commuting and vice versa, where
two projectors $P_{0}$ and~$P_{1}$ are \emph{commuting} if $P_{0}P_{1}=P_{1}P_{0}$.
Moreover, the events $E_{0}$ and $E_{1}$ in the same sample space
also satisfy the following properties: 
\begin{align}
\varphi\left(E_{0}\cap E_{1}\right) & =\varphi\left(E_{0}\right)\varphi\left(E_{1}\right)\,, & \varphi\left(E_{0}\cup E_{1}\right) & =\varphi\left(E_{0}\right)+\varphi\left(E_{1}\right)-\varphi\left(E_{0}\right)\varphi\left(E_{1}\right)\,.\label{eq:pullback-properties}
\end{align}
If $E_{0}$ and $E_{1}$ are disjoint, then $\varphi\left(E_{0}\right)$
and $\varphi\left(E_{1}\right)$ are orthogonal and vice versa, where
two projectors $P_{0}$ and $P_{1}$ are called \emph{orthogonal}
if $P_{0}P_{1}=\mathbb{0}$.
